---
title: "Reproducible Research: Peer Assessment 1"
date: "`r Sys.Date()`" 
author: Dr. M. Teunis
output: 
  html_document:
    keep_md: true
---

```{r, rprojroot}
# Project root
# Package {rprojroot} handles issues with the root in 
# knitting Rmd files.
## install.packages("rprojroot")
library(rprojroot)
```

```{r, knitr_setup}
# Knitr options and defining the root of the project
knitr::opts_chunk$set(echo = TRUE)
ROOT <- find_root_file(criterion = is_rstudio_project)
knitr::opts_knit$set(root.dir = ROOT)
```

```{r, packages, message=FALSE, warning=FALSE}
## install.packages("dplyr", dependencies = TRUE)
## install.packages("readr")
## install.packages("ggplot2")
## install.packages("pander")

library(dplyr)
library(readr)
library(ggplot2)
library(pander)

```

## Loading and preprocessing the data

### Unzipping the activity.zip file
```{r, unzip_data}

data_file <- paste0(ROOT, "/activity.zip") 
unzip(zipfile = data_file, exdir = "./data" )
``` 

### Reading the datafile
For reading, cleaning and subsetting the data I use the 'Hadley' verse packages: {readr}, {dplyr}.
```{r, reading_data}
activity <- read_csv(file = "./data/activity.csv")
``` 

### Removing NAs
```{r, na_omit}
activity <- na.omit(activity)
```

### Inspecting the activity dataframe
```{r, inspect_data}
head(activity)
glimpse(activity)
str(activity)
```


## What is mean total number of steps taken per day?

### Subsetting and summarize data
Converting $date to character, than to factor
```{r}
activity$date <- as.character(activity$date) 
activity$date <- as.factor(activity$date)
str(activity)
```

Checking the levels of $date.
```{r}
levels(activity$date)
``` 

Subsetting and calculating steps per day, using the pipe (%>%) operator. 
```{r}
library(dplyr)
activity$steps <- as.numeric(activity$steps)
tot_steps_pd <- activity %>%
            group_by(date) %>% 
            summarise(total_steps = sum(steps))
```

### Plotting total steps per day
```{r, fig.height=4, fig.width=12}
library(ggplot2)
# function to save ggplot2 plots
# this function takes the argument imageDirectory and filename as arguments

str(tot_steps_pd)
tot_steps_pd$date <- as.factor(tot_steps_pd$date)
names(tot_steps_pd)

# creating image directory
dir.create(paste0(ROOT, "/images"))

imageDirectory <- paste0(ROOT, "./images")
saveInImageDirectory<-function(imageDirectory,filename){
  imageFile <- file.path(imageDirectory, filename)
  ggsave(imageFile, dpi = 300, width = 8, height = 6)	
}

# plotting total steps per day
p <- ggplot(data = tot_steps_pd, aes(date, total_steps))
graph <- p + geom_point() +
  theme(plot.margin = unit(c(1,1,1,1), "cm")) +
      theme(axis.text.x = element_text(size=11, angle=90)) +
            labs(title = "Total steps per day") +
                  xlab("Date") + 
                  ylab("Total Steps per Day")

graph
``` 

```{r}
## saving the graph in "./images"
saveInImageDirectory(imageDirectory = imageDirectory,
filename = "mean_steps_per_day.png")
```

### Calculating and report mean and median of total number of steps per day
```{r}


activity$steps <- as.numeric(activity$steps)
activity$interval <- as.numeric(activity$interval)

mean_median_steps_pd <- activity %>%
            group_by(date) %>% 
            summarise(mean_steps = mean(steps), 
                      median_steps = median(steps))

overall_mean <- mean(mean_median_steps_pd$mean_steps)
overall_median <- mean(mean_median_steps_pd$median_steps)

pander(cbind(overall_mean, overall_median))

```


## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
